---
title: "POINT-seq Splicing Analysis Report"
output:
  pdf_document:
    toc: true
    number_sections: true
params:
  output_dir: "."
  generated_at: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

safe_head <- function(x, n = 10) {
  if (is.null(x) || nrow(x) == 0) {
    return(data.frame(note = "No rows available"))
  }
  utils::head(x, n)
}

sig_summary <- function(x, label) {
  if (is.null(x) || !"padj" %in% names(x)) {
    return(data.frame(analysis = label, tested_features = NA_integer_, padj_lt_0_05 = NA_integer_))
  }
  data.frame(
    analysis = label,
    tested_features = nrow(x),
    padj_lt_0_05 = sum(!is.na(x$padj) & x$padj < 0.05),
    stringsAsFactors = FALSE
  )
}

condition_counts <- as.data.frame(table(sample_table$condition), stringsAsFactors = FALSE)
names(condition_counts) <- c("condition", "sample_count")
```

# Run Summary

Generated: `r params$generated_at`

Project: `r cfg$project$name`

Genome build: `r cfg$project$genome_build`

Annotation: `r cfg$project$annotation_gtf`

Output directory: `r params$output_dir`

# Input Samples

```{r input-samples}
knitr::kable(sample_table, caption = "Included samples and BAM files")
```

```{r sample-counts}
knitr::kable(condition_counts, caption = "Number of included samples per condition")
```

# Input Data Statistics

```{r count-statistics}
count_stats <- data.frame(
  sample_id = sample_table$sample_id,
  gene_counts = colSums(DESeq2::counts(results$gene$dds)[, sample_table$sample_id, drop = FALSE]),
  intron_counts = colSums(DESeq2::counts(results$intron_retention$dds)[, sample_table$sample_id, drop = FALSE]),
  exon_bin_counts = colSums(DESeq2::counts(results$alternative_splicing$dds)[, sample_table$sample_id, drop = FALSE]),
  stringsAsFactors = FALSE
)
knitr::kable(count_stats, caption = "Assigned feature counts per sample")
```

```{r count-plot, fig.height=4, fig.width=7}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  long_counts <- rbind(
    data.frame(sample_id = count_stats$sample_id, feature_type = "Gene", assigned_counts = count_stats$gene_counts),
    data.frame(sample_id = count_stats$sample_id, feature_type = "Intron", assigned_counts = count_stats$intron_counts),
    data.frame(sample_id = count_stats$sample_id, feature_type = "Exon bin", assigned_counts = count_stats$exon_bin_counts)
  )
  ggplot2::ggplot(long_counts, ggplot2::aes(x = sample_id, y = assigned_counts, fill = feature_type)) +
    ggplot2::geom_col(position = "dodge") +
    ggplot2::theme_bw() +
    ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1)) +
    ggplot2::labs(x = "Sample", y = "Assigned counts", fill = "Feature type")
}
```

# Key Results Summary

```{r result-summary}
summary_table <- rbind(
  sig_summary(results$gene$table, "Differential nascent transcription"),
  sig_summary(results$intron_retention$table, "Differential intron retention"),
  sig_summary(results$alternative_splicing$table, "Alternative splicing exon usage"),
  sig_summary(results$splicing_efficiency$table, "Differential splicing efficiency")
)
knitr::kable(summary_table, caption = "Significant feature summary")
```

# Differential Nascent Transcription

```{r top-genes}
knitr::kable(
  safe_head(results$gene$table[, intersect(c("feature_id", "gene_id", "seqnames", "start", "end", "log2FoldChange", "pvalue", "padj"), names(results$gene$table)), drop = FALSE]),
  caption = "Top differential nascent transcription results"
)
```

# Normalized Intron Coverage

```{r top-normalized-intron}
knitr::kable(
  safe_head(results$normalized_intron$table[, intersect(c("feature_id", "gene_id", "seqnames", "start", "end", "log2_fold_change"), names(results$normalized_intron$table)), drop = FALSE]),
  caption = "Top normalized intron coverage changes"
)
```

# Differential Intron Retention

```{r top-ir}
knitr::kable(
  safe_head(results$intron_retention$table[, intersect(c("feature_id", "gene_id", "seqnames", "start", "end", "log2FoldChange", "pvalue", "padj"), names(results$intron_retention$table)), drop = FALSE]),
  caption = "Top differential intron retention results"
)
```

# Alternative Splicing

```{r top-as}
knitr::kable(
  safe_head(results$alternative_splicing$table[, intersect(c("feature_id", "gene_id", "seqnames", "start", "end", "log2FoldChange", "pvalue", "padj"), names(results$alternative_splicing$table)), drop = FALSE]),
  caption = "Top exon-bin alternative splicing proxy results"
)
```

# Splicing Efficiency

```{r top-se}
knitr::kable(
  safe_head(results$splicing_efficiency$table[, intersect(c("feature_id", "gene_id", "seqnames", "start", "end", "log2FoldChange", "pvalue", "padj"), names(results$splicing_efficiency$table)), drop = FALSE]),
  caption = "Top differential splicing efficiency results"
)
```

# Genome Browser Outputs

The pipeline writes BED, bedGraph, and attempted BigWig outputs under:

```{r browser-files}
browser_dir <- file.path(params$output_dir, "browser")
browser_files <- if (dir.exists(browser_dir)) {
  data.frame(file = list.files(browser_dir, recursive = TRUE), stringsAsFactors = FALSE)
} else {
  data.frame(file = "No browser output directory found", stringsAsFactors = FALSE)
}
knitr::kable(browser_files, caption = "Genome browser output files")
```

# Notes

The first implementation uses exon-bin differential usage as an alternative splicing proxy. Event-specific alternative splicing calls can be added after validating the behavior on pilot POINT-seq data.
